Forecasting the US Election 2024

### 2024-11-01 paper/paper.qmd
update paper and citations

---
title: "Forcasting US Election 2024"
subtitle: "##TODO: result here"
author: 
  - Maryam Ansari
  - Amy Jin
  - Maggie Zhang

@@

As of October-November 2024, the U.S. presidential election is almost at its final stage, with all candidates actively campaigning across numerous states. The presidential election will happen on November 5, 2024 [@calendar]. The two major parties, Democrats and Republicans, are in particular focus. Former President Donald Trump represents the Republican party, and President Joe Biden initially led re-election for the Democrats. However, Biden has dropped out, and Kamala Harris has taken over as the Democratic nominee [@g24], who is the first Black woman nominee. Harris is the Both candidates are focusing their efforts on securing votes by addressing their aspect on social issues. pollsters and election forecasting with data evidence are important in predicting election outcomes. Polling can highlight underlying or emerging issues, as well as reveal voter preferences. The polling outcomes allow campaigns of different parties to strategically target specific demographics. For example in 2012, pollsters and data modellers such as Nate Silver have all used survey research and statistical models to successfully predict the result of Barack Obama's victory in the presidential election with consistent and reliable forecasts [@b17].

@@

## Overview

@@

The dataset comprises several key variables of interest, including poll_id, a unique identifier for each poll conducted; pollster, which indicates the organization conducting the poll; sample_size, representing the total number of respondents; population, specifying the voting group described (e.g., likely voters); candidate_name, the names of the candidates in the poll (e.g., Kamala Harris, Donald Trump); and pct, the percentage of the vote or support received by each candidate. These variables allow us to explore various dimensions of polling data, such as trends in voter support across different states and the influence of pollster reliability on polling outcomes. The [@tbl-firstten] below provides a snapshot of the polling data, displaying the first ten entries. This includes the pollster names, sample sizes, and the percentage of support for each candidate, which can reveal patterns in public opinion and help identify how different organizations may report varying levels of support for Kamala Harris and Donald Trump.

```{r}
#| label: tbl-firstten
#| tbl-cap: "Sample sizes and support percentages for Kamala Harris and Donald Trump"
#| echo: false
# Drop rows with NA values and select only the relevant columns
firstten_data <- drop_na(data) %>%
  select(pollster_id, pollster, sample_size, population, candidate_name, pct)

@@

# Create the gt table with only the selected columns
firstten <- first_ten_entries %>%
  gt() %>%
  tab_header(
    title = "Polling Data Snapshot: 2024 U.S. Presidential Election",
  ) %>%
  cols_label(
    pollster_id = "Poll ID",
    pollster = "Pollster",
    sample_size = "Sample Size",
    population = "Population",
    candidate_name = "Candidate",
    pct = "Support (%)"
  ) %>%
  fmt_number(
    columns = c(sample_size, pct),
    decimals = 0
  )

@@

@@

The relationship between polling organization reliability, as indicated by the pollscore, and the percentage of support (pct) for each candidate was examined. It is anticipated that more reliable pollsters will yield more accurate predictions. Consequently, polls were categorized into three tiers based on their pollscore: high reliability (pollscore greater than 0), medium reliability (pollscore between -1 and 0), and low reliability (pollscore less than -1). By comparing the average support percentages (pct) for Kamala Harris and Donald Trump across these tiers, this analysis sought to identify whether more reliable pollsters produce different outcomes than those deemed less reliable. Mean pct values were calculated for each category, and standard deviation was included to illustrate variability within the poll results. The table[@tbl-reliability] and bar chart[@fig-reliability] showing the average `pct` for each candidate across different levels of pollster reliability. Discussion: This analysis highlights whether pollsters with higher reliability scores offer more accurate predictions and if their estimates favor one candidate over the other.

```{r}
#| label: tbl-reliability
#| tbl-cap: "Mean and Standard Deviation of Support Percentages (pct) for Kamala Harris and Donald Trump Across Pollster Reliability Levels"
#| echo: false
# Categorizing poll reliability
# Create a new dataset with categorized reliability
# Remove rows with NA values in the 'pollscore' column while keeping the rest of the data frame
pollscore_data <- data %>%
  filter(!is.na(pollscore)) %>%
  mutate(
    reliability = case_when(
      pollscore > 0 ~ "High",
      pollscore > -1 & pollscore <= 0 ~ "Medium",
      pollscore <= -1 ~ "Low"
    )
  )

# Calculate mean and standard deviation for each candidate across reliability levels
summary_data <- pollscore_data %>%
  group_by(candidate_name, reliability) %>%
  summarise(
    mean_pct = mean(pct),
    sd_pct = sd(pct),
    .groups = 'drop'
  )
reliability_table <- summary_data %>%
  gt() %>%
  tab_header(
    title = "Polling Data Summary: Pollster Reliability and Support Percentages"
  ) %>%
  cols_label(
    candidate_name = "Candidate",
    reliability = "Pollster Reliability",
    mean_pct = "Mean Support (%)",
    sd_pct = "Standard Deviation"
  ) %>%
  fmt_number(
    columns = c(mean_pct, sd_pct),
    decimals = 2
  )

@@

```{r}
#| label: fig-reliability
#| fig-cap: "Graph of Dates Categorized by Year and Presented by Each Months"
#| echo: false
library(ggplot2)

# Create a more aesthetic bar chart with alpha and no labels
ggplot(summary_data, aes(x = reliability, y = mean_pct, fill = candidate_name)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7, color = "black", alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_pct - sd_pct, ymax = mean_pct + sd_pct), 
                width = 0.2, 
                position = position_dodge(0.7)) +
  labs(
    title = "Average Support Percentage by Pollster Reliability",
    subtitle = "Comparison of support percentages for Kamala Harris and Donald Trump",
    x = "Pollster Reliability",
    y = "Average Support (%)",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = "#1f78b4", "Donald Trump" = "#e31a1c")) + # Custom colors
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title.x = element_text(face = "bold", margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", margin = margin(r = 10)),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )

@@

# Create a table using the gt package
reliability_table <- categorized_data %>%
  gt() %>%
  tab_header(
    title = "Polling Data Summary: Methodology and Support Percentages",
    subtitle = "Mean Support Percentages (pct) for Kamala Harris and Donald Trump by Polling Methodology Category"
  ) %>%
  cols_label(
    methodology = "Polling Methodology",
    `Donald Trump Mean Support (%)` = "Trump(%)",
    `Kamala Harris Mean Support (%)` = "Harris(%)",
    Difference = "Difference"
  ) %>%
  fmt_number(
    columns = c(`Donald Trump Mean Support (%)`, `Kamala Harris Mean Support (%)`, Difference),
    decimals = 2
  )
# Display the table
reliability_table

### State-by-State Polling Trends

Given the importance of state-level polling data in U.S. presidential elections, this analysis focused on key battleground states (e.g., Florida, Pennsylvania, Wisconsin). The polling results for Kamala Harris and Donald Trump were examined by state (state), and the average pct was calculated for each swing state. These trends were visualized to assess whether one candidate consistently received more support in these critical states, as such patterns could significantly influence the final election outcome. A state-by-state breakdown of the polling percentages for each candidate, with a focus on swing states is shown by [@fig-state]. Discussion: The results highlight key states where the election might be decided, helping to pinpoint potential trends favoring one candidate.

```{r}
#| label: fig-state
#| fig-cap: "Graph of Dates Categorized by Year and Presented by Each Months"
#| echo: false
# Calculate average pct by state and candidate
library(tidyverse)
library(ggplot2)

# Filter out rows with NA values in the 'state' column and keep only rows for Kamala Harris and Donald Trump
filtered_data <- data %>%
  filter(!is.na(state)) %>%
  filter(candidate_name %in% c("Kamala Harris", "Donald Trump"))

# Calculate the average support percentage for each candidate by state
state_summary_data <- filtered_data %>%
  group_by(state, candidate_name) %>%
  summarise(
    avg_pct = mean(pct),
    .groups = 'drop'
  )

# Create a state-by-state breakdown bar plot of polling percentages
ggplot(state_summary_data, aes(x = state, y = avg_pct, fill = candidate_name)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7) +
  labs(
    title = "State-by-State Polling Trends in Key Battleground States",
    subtitle = "Average support percentages for Kamala Harris and Donald Trump",
    x = "State",
    y = "Average Support Percentage (%)",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = "#1f78b4", "Donald Trump" = "#e31a1c")) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title.x = element_text(face = "bold", margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", margin = margin(r = 10)),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )

@@

```{r}
#| label: fig-samplesize
#| fig-cap: "Graph of Dates Categorized by Year and Presented by Each Months"
#| echo: false
# Categorizing sample size into small, medium, and large groups
library(tidyverse)
library(ggplot2)

@@

# Create the chart comparing candidate vote percentages by sample size category
ggplot(summary_data_sample_size, aes(x = sample_size_category, y = avg_pct, fill = candidate_name)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7, color = "black", alpha = 0.8) +
  geom_errorbar(aes(ymin = avg_pct - sd_pct, ymax = avg_pct + sd_pct), 
                width = 0.2, 
                position = position_dodge(0.7)) +
  labs(
    title = "Candidate Vote Percentages by Sample Size Category",
    subtitle = "Average support percentages for Kamala Harris and Donald Trump based on sample size groups",
    x = "Sample Size Category",
    y = "Average Support Percentage (%)",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = "#1f78b4", "Donald Trump" = "#e31a1c")) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title.x = element_text(face = "bold", margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", margin = margin(r = 10)),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )

@@

Alternate Analysis Data 5: Sponsorship and Bias in Polling Finally, the potential bias of polls sponsored by partisan organizations was assessed to determine whether such sponsorship influenced support for the associated candidate. The vote percentages (pct) for Kamala Harris and Donald Trump were compared in polls sponsored by Democratic-leaning and Republican-leaning organizations (sponsor_candidate_party). By calculating the average pct for each candidate within partisan-sponsored polls, this analysis explored whether partisan sponsorship resulted in a systematic overestimation of support for a particular candidate. A comparison of the polling results in partisan-sponsored polls, broken down by party affiliation is shown by [@fig-sponsorship]. Discussion: This analysis reveals potential biases in partisan-sponsored polls and assesses the objectivity of different polling organizations.

```{r}
#| label: fig-sponsorship
#| fig-cap: "Graph of Dates Categorized by Year and Presented by Each Months"
#| echo: false
# Calculate the average pct and standard deviation for each candidate by sponsorship type
sponsorship_summary_data <- data %>%
  filter(sponsor_candidate_party != "Non-Partisan") %>%  # Focus on partisan-sponsored polls
  group_by(candidate_name, sponsor_candidate_party) %>%
  summarise(
    avg_pct = mean(pct),
    sd_pct = sd(pct),
    .groups = 'drop'
  )
library(ggplot2)

# Create the bar chart comparing vote percentages by sponsorship type
ggplot(sponsorship_summary_data, aes(x = sponsor_candidate_party, y = avg_pct, fill = candidate_name)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7, color = "black", alpha = 0.8) +
  geom_errorbar(aes(ymin = avg_pct - sd_pct, ymax = avg_pct + sd_pct), 
                width = 0.2, 
                position = position_dodge(0.7)) +
  labs(
    title = "Vote Percentages by Partisan Sponsorship",
    subtitle = "Average support percentages for Kamala Harris and Donald Trump in partisan-sponsored polls",
    x = "Sponsorship (Party Affiliation)",
    y = "Average Support Percentage (%)",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = "#1f78b4", "Donald Trump" = "#e31a1c")) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title.x = element_text(face = "bold", margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", margin = margin(r = 10)),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )

@@

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset. A thorough discussion of measurement, relating to the dataset, is provided in the data section. Please ensure that you explain how we went from some phenomena in the world that happened to an entry in the dataset that you are interested in.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins [@fig-bills], from @palmerpenguins.

Talk more about it.

And also planes [@fig-planes]. (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

The model should be nicely written out, well-explained, justified, and appropriate.

Detail the statistical model used, defining and explaining each aspect and its importance, and ensure that variables are well-defined and correspond with those discussed in the data section.

The model should have an appropriate balance of complexity—neither overly simplistic nor unnecessarily complicated—and be justified as suitable for the situation.

## Modeling decisions

Explain how decisions made in modeling reflect the aspects discussed in the data section, including why specific features are included (e.g., why use age rather than age-groups, treating province effects as levels, categorizing gender, etc?).

## Mathematical notations

Present the model using appropriate mathematical notation supplemented with plain English explanations, defining every component. If applicable, define sensible priors for Bayesian models.

## Assumptions, limitations, circumstances

Clearly discuss the underlying assumptions, potential limitations, and circumstances where the model may not be appropriate.

## Software & model validation

Mention the software used to implement the model, and provide evidence of model validation and checking—such as out-of-sample testing, RMSE calculations, test/training splits, or sensitivity analyses—while addressing model convergence, diagnostics, and any alternative models or variants considered,

## Strenth and weeknesses and final choice

including their strengths and weaknesses and the rationale for the final model choice.

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@talia] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

>>>>

---
title: "Forecasting the US Election 2024"
subtitle: "##TODO: result here"
author: 
  - Maryam Ansari
  - Amy Jin
  - Maggie Zhang

@@

As of October-November 2024, the U.S. presidential election is almost at its final stage, with all candidates actively campaigning across numerous states. The presidential election will happen on November 5, 2024 [@calendar]. The two major parties, Democrats and Republicans, are in particular focus. Former President Donald Trump represents the Republican party, and President Joe Biden initially led re-election for the Democrats. However, Biden has dropped out, and Kamala Harris has taken over as the Democratic nominee [@g24], who is the first Black woman nominee. Harris is the Both candidates are focusing their efforts on securing votes by addressing their aspect on social issues.

Pollsters and election forecasting with data evidence are important in predicting election outcomes. Polling can highlight underlying or emerging issues, as well as reveal voter preferences. The polling outcomes allow campaigns of different parties to strategically target specific demographics. For example in 2012, pollsters and data modellers such as Nate Silver have all used survey research and statistical models to successfully predict the result of Barack Obama's victory in the presidential election with consistent and reliable forecasts [@b17].

@@

## Overview {#sec-dataoverview}

@@

The dataset comprises several key variables of interest, including but not limited to,

**poll_id:** a unique identifier for each poll conducted pollster: which indicates the organization conducting the poll

**sample_size:** representing the total number of respondents

**population:** specifying the voting group described (e.g., likely voters)

**candidate_name:** the names of the candidates in the poll (e.g., Kamala Harris, Donald Trump)

**pct:** the percentage of the vote or support received by each candidate.

These variables allow us to explore various dimensions of polling data, such as trends in voter support across different states and the influence of pollster reliability on polling outcomes. The table below provides a snapshot of the polling data, displaying the first ten entries. This includes the pollster names, sample sizes, and the percentage of support for each candidate, which can reveal patterns in public opinion and help identify how different organizations may report varying levels of support for Kamala Harris and Donald Trump.

### Response variables:

The main response variable of our analysis is the percentage support (pct) for Donald Trump and Kamala Harris shown in each poll. This number tells us what percentage of people said they would vote for each candidate when the survey was taken. We use it as the basis of our electoral prediction as it helps us understand how people feel about the candidates.

TODO: descibe

```{r}
#| label: tbl-pct
#| tbl-cap: "Histogram of Trump Support Percentage (pct)"
#| echo: false
ggplot(data, aes(x = pct)) +
  geom_histogram(binwidth = 5, fill = "#e31a1c", color = "black") +
  theme_minimal() +
  labs(
    x = "Percentage Support (pct)",
    y = "Frequency"
  )
```

### Predictor variables:

Our model takes into account various variables; their names and descriptions are,

**Pollster**: The name of the polling organization that conducted the poll (e.g., YouGov, RMG Research).

**State:** The U.S. state where the poll was conducted.

**Sample Size:** The total number of respondents that participated in the poll. End Date: The date the poll ended.

These variables were selected and kept due to their evident relationship with the response variable and the significance of it. More details of the relationships and model are provided in later sections.

```{r}
#| label: tbl-firstten
#| tbl-cap: "Polling Data Snapshot: 2024 U.S. Presidential Election"
#| echo: false
# Drop rows with NA values and select only the relevant columns
firstten_data <- drop_na(data) %>%
  select(pollster_id, pollster, sample_size, population, candidate_name, pct)

@@

# Create the gt table with only the selected columns
firstten <- first_ten_entries %>%
  gt() %>%
  cols_label(
    pollster_id = "Poll ID",
    pollster = "Pollster",
    sample_size = "Sample Size",
    population = "Population",
    candidate_name = "Candidate",
    pct = "Support (%)"
  ) %>%
  fmt_number(
    columns = c(sample_size, pct),
    decimals = 0
  )

@@

### Data Manipulation and cleaning:

The data was cleaned, filtered and mutated to better align with our prediction model. The software and packages used have been highlighted in @sec-dataoverview. The key steps of our cleaning process were:

Filtered out lower quality polls by only keeping polls that met a set numeric grade threshold (numeric_grade \>= 2.5). Filtered out pollsters that had less than 5 entries. Filtered out data that was collected before July 21st, 2024. Added a variable num_trump to count the number of votes for Donald Trump using pct. For data entries where the state was missing, they were categorised as “National”, to standardize the data.

The dataset had several other variables that we did not include in our analysis. These variables were either majorly missing values or not informative enough to be included.

@@

The relationship between polling organization reliability, as indicated by the pollscore, and the percentage of support (pct) for each candidate was examined. It is anticipated that more reliable pollsters will yield more accurate predictions. Consequently, polls were categorized into three tiers based on their pollscore: high reliability (pollscore greater than 0), medium reliability (pollscore between -1 and 0), and low reliability (pollscore less than -1). By comparing the average support percentages (pct) for Kamala Harris and Donald Trump across these tiers, this analysis sought to identify whether more reliable pollsters produce different outcomes than those deemed less reliable. Mean pct values were calculated for each category, and standard deviation was included to illustrate variability within the poll results. The table (@tbl-reliability) and bar chart @fig-reliability showing the average `pct` for each candidate across different levels of pollster reliability. Discussion: This analysis highlights whether pollsters with higher reliability scores offer more accurate predictions and if their estimates favor one candidate over the other.

```{r}
#| label: tbl-reliability
#| tbl-cap: "Summary Statistics for Support Percentages Across Pollster Reliability Levels"
#| echo: false
# Categorizing poll reliability
# Create a new dataset with categorized reliability
# Remove rows with NA values in the 'pollscore' column while keeping the rest of the data frame
pollscore_data <- data %>%
  filter(!is.na(pollscore)) %>%
  mutate(
    reliability = case_when(
      pollscore > 0 ~ "High",
      pollscore > -1 & pollscore <= 0 ~ "Medium",
      pollscore <= -1 ~ "Low"
    )
  )

# Calculate mean and standard deviation for each candidate across reliability levels
summary_data <- pollscore_data %>%
  group_by(candidate_name, reliability) %>%
  summarise(
    mean_pct = mean(pct),
    sd_pct = sd(pct),
    .groups = 'drop'
  )
reliability_table <- summary_data %>%
  gt() %>%
  cols_label(
    candidate_name = "Candidate",
    reliability = "Pollster Reliability",
    mean_pct = "Mean Support (%)",
    sd_pct = "Standard Deviation"
  ) %>%
  fmt_number(
    columns = c(mean_pct, sd_pct),
    decimals = 2
  )

@@

```{r}
#| label: fig-reliability
#| fig-cap: "Average Support Percentage by Pollster Reliability"
#| echo: false
library(ggplot2)

# Create a more aesthetic bar chart with alpha and no labels
ggplot(summary_data, aes(x = reliability, y = mean_pct, fill = candidate_name)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7, color = "black", alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_pct - sd_pct, ymax = mean_pct + sd_pct), 
                width = 0.2, 
                position = position_dodge(0.7)) +
  labs(
    subtitle = "Comparison of support percentages for Kamala Harris and Donald Trump",
    x = "Pollster Reliability",
    y = "Average Support (%)",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = "#1f78b4", "Donald Trump" = "#e31a1c")) + # Custom colors
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title.x = element_text(face = "bold", margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", margin = margin(r = 10)),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )

@@

# Create a table using the gt package
reliability_table <- categorized_data %>%
  gt() %>%
  cols_label(
    methodology = "Polling Methodology",
    `Donald Trump Mean Support (%)` = "Trump(%)",
    `Kamala Harris Mean Support (%)` = "Harris(%)",
    Difference = "Difference"
  ) %>%
  fmt_number(
    columns = c(`Donald Trump Mean Support (%)`, `Kamala Harris Mean Support (%)`, Difference),
    decimals = 2
  )
# Display the table
reliability_table

@@

```{r}
#| label: fig-samplesize
#| fig-cap: "Candidate Vote Percentages by Sample Size Category"
#| echo: false
# Categorizing sample size into small, medium, and large groups
library(tidyverse)
library(ggplot2)

@@

# Create the chart comparing candidate vote percentages by sample size category
ggplot(summary_data_sample_size, aes(x = sample_size_category, y = avg_pct, fill = candidate_name)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7, color = "black", alpha = 0.8) +
  geom_errorbar(aes(ymin = avg_pct - sd_pct, ymax = avg_pct + sd_pct), 
                width = 0.2, 
                position = position_dodge(0.7)) +
  labs(
    x = "Sample Size Category",
    y = "Average Support Percentage (%)",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = "#1f78b4", "Donald Trump" = "#e31a1c")) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title.x = element_text(face = "bold", margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", margin = margin(r = 10)),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )

@@

### Sponsorship and Bias in Polling

Finally, the potential bias of polls sponsored by partisan organizations was assessed to determine whether such sponsorship influenced support for the associated candidate. The vote percentages (pct) for Kamala Harris and Donald Trump were compared in polls sponsored by Democratic-leaning and Republican-leaning organizations (sponsor_candidate_party). By calculating the average pct for each candidate within partisan-sponsored polls, this analysis explored whether partisan sponsorship resulted in a systematic overestimation of support for a particular candidate. A comparison of the polling results in partisan-sponsored polls, broken down by party affiliation is shown by [@fig-sponsorship]. Discussion: This analysis reveals potential biases in partisan-sponsored polls and assesses the objectivity of different polling organizations.

```{r}
#| label: fig-sponsorship
#| fig-cap: "Vote Percentages by Partisan Sponsorship"
#| echo: false
# Calculate the average pct and standard deviation for each candidate by sponsorship type
sponsorship_summary_data <- data %>%
  filter(sponsor_candidate_party != "Non-Partisan") %>%  # Focus on partisan-sponsored polls
  group_by(candidate_name, sponsor_candidate_party) %>%
  summarise(
    avg_pct = mean(pct),
    sd_pct = sd(pct),
    .groups = 'drop'
  )
library(ggplot2)

# Create the bar chart comparing vote percentages by sponsorship type
ggplot(sponsorship_summary_data, aes(x = sponsor_candidate_party, y = avg_pct, fill = candidate_name)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7, color = "black", alpha = 0.8) +
  geom_errorbar(aes(ymin = avg_pct - sd_pct, ymax = avg_pct + sd_pct), 
                width = 0.2, 
                position = position_dodge(0.7)) +
  labs(
    x = "Sponsorship (Party Affiliation)",
    y = "Average Support Percentage (%)",
    fill = "Candidate"
  ) +
  scale_fill_manual(values = c("Kamala Harris" = "#1f78b4", "Donald Trump" = "#e31a1c")) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.title.x = element_text(face = "bold", margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", margin = margin(r = 10)),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )

@@

## Measurement and Limitations

This dataset comes with several important measurement and limitations:

**Polling Accuracy:** While efforts are made to prioritize high-quality polls through numeric ratings, differences in polling methods can still introduce some degree of bias. Pollster ratings enhance reliability but cannot fully eliminate methodological inconsistencies.

**Time Sensitivity:** The dataset captures voter preferences at particular points rather than over time, providing only a static view. This limitation means that shifts in voter sentiment throughout the campaign are not represented.

**Geographical Coverage:** The dataset includes both national and state-level polling, but coverage is uneven. High-interest states, like swing states, are polled more frequently, while data from other regions may be sparse, potentially affecting regional analysis.

**Participation Bias**: Even with strict rules for polling, certain biases remain possible, such as selection bias in survey participation and response bias due to social desirability factors.

# Statistical Model and Analysis

## Model Overview

The goal of our modeling is to predict which candidate from Donald Trump and Kamala Harris will win the 2024 US election. We will firstly, predict support for the two candidates using the tailored cleaned data in the 2024 US presidential election using aggregated polling data, and secondly, to select a model that balances complexity and interpretability while accurately capturing the key trends and effects. This section describes the statistical models used, justifies the modeling choices, and discusses the assumptions, limitations, and validation strategies. Background details, including diagnostics and additional model exploration, are available in [Appendix -@sec-model-details].

We utilized both traditional linear models and Bayesian models. The Bayesian approach, in particular, allows us to incorporate prior beliefs and better quantify uncertainty, which is crucial given the inherent variability in polling data.

#### Model Specifications

We constructed four models of increasing complexity:

**Model 1**: A linear regression where the dependent variable,

pct, the percentage of respondents supporting each candidate, is modeled as a function of the date the poll ended,

end_date, to capture temporal trends.

**Model 2**: An extension of Model 1 that includes a categorical variable,

pollster, to account for systematic differences between polling organizations.

**Model 3**: Adds a binary indicator,

is_national, to Model 2 to differentiate between national and state-specific polls.

**Model 4**: A Bayesian model implemented using the `rstanarm` package, incorporating the same predictors as Model 3 but allowing for more robust estimation under uncertainty.

### We mathematically define these models as follows:

#### Model 1

$$
y_i = \beta_0 + \beta_1 \cdot \text{end\_date}_i + \epsilon_i, \quad \epsilon_i \sim \text{Normal}(0, \sigma^2)
$$

#### Model 2

$$
y_i = \beta_0 + \beta_1 \cdot \text{end\_date}_i + \sum_{j=1}^J \gamma_j \cdot \text{pollster}_{ij} + \epsilon_i, \quad \epsilon_i \sim \text{Normal}(0, \sigma^2)
$$

#### Model 3

$$
y_i = \beta_0 + \beta_1 \cdot \text{end\_date}_i + \sum_{j=1}^J \gamma_j \cdot \text{pollster}_{ij} + \delta \cdot \text{is\_national}_i + \epsilon_i, \quad \epsilon_i \sim \text{Normal}(0, \sigma^2)
$$

#### Model 4 (Bayesian)

$$
y_i \mid \mu_i, \sigma \sim \text{Normal}(\mu_i, \sigma), \quad \mu_i = \beta_0 + \beta_1 \cdot \text{end\_date}_i + \sum_{j=1}^J \gamma_j \cdot \text{pollster}_{ij} + \delta \cdot \text{is\_national}_i
$$

**Priors for Model 4:**

$$
\beta_0, \beta_1, \gamma_j, \delta \sim \text{Normal}(0, 5), \quad \sigma \sim \text{Exponential}(1)
$$

#### Variables and Justification

1.  **Dependent Variable**: ( y_i ) (support percentage, `pct`).
2.  **Predictors**:
    -   **(** \text{end\_date} ): Captures time trends, which are essential in understanding shifts in public opinion.
    -   **(** \text{pollster} ): Accounts for differences in polling methodologies and potential biases.
    -   **(** \text{is\_national} ): Differentiates between polls conducted at the national level and those specific to individual states, as national polls tend to exhibit different characteristics and sample compositions compared to state polls.

These modeling choices reflect insights from the data section, where we observed systematic variations in support percentages over time, across pollsters, and between national and state polls.

#### Model Justification

The modeling approach was designed to balance interpretability and predictive power:

**Model 1** serves as a baseline, capturing the temporal trend.

**Model 2** addresses variability between pollsters, which is crucial given differences in methodologies and sample recruitment.

**Model 3** adds another layer of nuance by accounting for the national vs. state-specific distinction.

**Model 4** employs Bayesian inference, allowing us to incorporate uncertainty and use prior distributions for more robust predictions.

The Bayesian approach is particularly useful given the variability in polling data. Using `rstanarm`, we employ default weakly informative priors, which are appropriate given the lack of strong prior information. These priors help regularize the estimates and prevent overfitting.

#### Assumptions, Limitations, and Circumstances

1.  **Assumptions**:
    -   Linear relationships between predictors and the outcome.
    -   Independence of observations, which may be violated if polls from the same organization are correlated.
    -   Normality of residuals in linear models.
2.  **Limitations**:
    -   The models do not account for potential interactions between variables or non-linear effects, which could improve predictive accuracy.
    -   The Bayesian model's convergence can be sensitive to the choice of priors and the data's variability.
3.  **Circumstances**:
    -   The models may be less appropriate in highly volatile electoral environments where sudden shifts in public opinion occur.
    -   The temporal model assumes a relatively smooth trend, which may not hold in the presence of significant events influencing voter sentiment.

#### Software and Model Validation

The models were implemented in R using the `tidyverse` and `rstanarm` packages. We assessed model performance using:

**AIC/BIC**: For model selection based on goodness-of-fit and penalization for complexity.

**Cross-validation**: (To be detailed further in the appendix) to evaluate out-of-sample performance.

**Residual Diagnostics**: To check for violations of model assumptions.

**Model Convergence**: Monitored using diagnostics provided by `rstanarm` for the Bayesian model.

#### Strengths, Weaknesses, and Final Model Choice

-   **Strengths**: The models are interpretable and straightforward, with the Bayesian approach providing a robust framework for uncertainty quantification.
-   **Weaknesses**: Potential oversimplification and assumptions of linearity. Future iterations could explore non-linear effects or hierarchical models.
-   **Final Choice**: We selected **Model 3** (linear model with `end_date`, `pollster`, and `is_national`) for its balance of interpretability and predictive performance. The Bayesian model serves as a robustness check, with results presented in the appendix.

Our final model selection and analysis provide a reliable and transparent prediction framework for Kamala Harris’s support in the 2024 US election.

@@

# Appendix B: Idealized Survey Methodology and Survey Design

## Methodology Overview

We will be developing a prediction model and conducting a survey targeting eligible voters in the United States for the 2024 presidential election. The study will observe key demographic groups influencing voting behaviour, such as age, ethnicity, income, education, and political affiliation [@leighley2013who]. Employing a combined stratified and quota sampling approach, we will ensure representation across these demographics, with a particular emphasis on swing states to enhance the accuracy of our predictions [@fitzgerald2024seven]. Recruitment will utilize both online and offline methods, including community outreach and random digit dialling, while incentives will be offered to boost participation. Data validation will involve screening questions and consistency checks to ensure reliable responses. Finally, we will implement hybrid modelling techniques to aggregate survey data and apply statistical models for more accurate electoral predictions, correcting for any biases through weighting methods [@pasek2015predicting] [@wlezien2002timeline]. A copy of our designed survey will be attached.

## Target Population

Target Population: The target population for our prediction model and survey includes all eligible voters across the United States participating in the upcoming 2024 presidential election. Population context study: We will observe specific key groups within the population that will affect people’s voting actions. These include age groups, racial and ethnic, educational levels geographic regions, political affiliation, etc. [@leighley2013who].

## Sampling Approach
